---
title: "Data Visualization Final Project Presentation"
author: "Joanna Wright"
institute: "Univeristy of Oregon"
date: "March 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

```

class: center, middle

# Exploring kindergarten readiness in Oregon   

---

# About the dataset

The data for this project are from the **Oregon Kindergarten Aseessment** and can be found [here](https://www.oregon.gov/ode/educator-resources/assessment/Pages/Kindergarten-Assessment.aspx). 

The OKA aims to provide *"a snapshot for communities and schools on the skills that students have when they enter Kindergarten."*

- Early Literacy
- Early Math
- Approaches to Learning (including self-regulation) 

*__"domains that are strongly linked to third grade reading and future academic success"__* 

---

# Visualization 1: mapping self-regulation by county 

```{r include=FALSE, echo=FALSE, results='hide'}

library(tidyverse)
library(rio)
library(here)
library(janitor)
library(maps)
library(mapdata)
library(ggtext)

d <- import(here("data", "ka_2018-19_lookback_report.xlsx")) %>%
  clean_names()

d_tidy <- d %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

d_tidy <- d_tidy %>%
  filter(avg_self_reg != "*")

d_tidy = d_tidy[-1,]

str(d_tidy)
d_tidy$avg_self_reg = as.numeric(d_tidy$avg_self_reg)

counties <- map_data("county")
or_county <- subset(counties, region == "oregon")

# identifying keys, joining: 

d %>%
  count(oregon_kindergarten_assessment, x2, x3, x4, x5, x6, x7, x8) %>%
  filter(n>1)

or_county %>%
  count(order) %>%
  filter(n>1)

d1 <- d_tidy %>%
  group_by(county) %>%
  summarise(avg_self_reg_county = mean(avg_self_reg))

or_county <- or_county %>%
  rename(county = subregion)

d1 <- d1 %>%
  mutate(county = str_to_lower(county))

join1 <- left_join(d1, or_county)

```

```{r echo=FALSE}

join1 %>%
ggplot() + 
    geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) + 
    coord_fixed(1.3) +
    scale_fill_viridis_c(name = "Score (1-5 scale)", 
                         option = "inferno") +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         caption = "Data from https://www.oregon.gov/ode/educator-resources/assessment/Pages/Kindergarten-Assessment.aspx",
         x = NULL,
         y = NULL) +
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank())

```

---

# Visualization 2: self-regulation scores by race

```{r include=FALSE, echo=FALSE, results='hide'}

d2 <- d_tidy %>%
  filter(stu_grp == "Asian" |
           stu_grp == "White" |
           stu_grp == "Multi-Racial" |
           stu_grp == "Native Hawaiian/Pacific Islander" |
           stu_grp == "Hispanic/Latino" |
           stu_grp == "Black/African American" |
           stu_grp == "American Indian/Alaska Native")


```

```{r echo=FALSE, warning=FALSE}

d2 %>%
  ggplot(aes(avg_self_reg, fct_reorder(stu_grp, avg_self_reg), fill = stu_grp)) +
  geom_violin() +
  labs(title = "Distribution of self-regulation scores by race",
       x = "Score (1-5 scale)",
       y = NULL) +
  theme_minimal() + 
  scale_color_brewer(palette = "Accent") +
  theme(plot.title = element_text(hjust = 1)) +
  theme(legend.position = "none") 

```

---
# Still Visualization 2: a closer look at differences by race

```{r echo=FALSE, warning=FALSE}

d2 %>%
  filter(stu_grp == "White" |
           stu_grp == "Hispanic/Latino") %>%
  group_by(stu_grp, county) %>%
  summarise(avg_self_reg_race = mean(avg_self_reg)) %>%
  pivot_wider(
    names_from = stu_grp,
    values_from = avg_self_reg_race
  ) %>%
  mutate(white_hisp_diff = c(White - `Hispanic/Latino`)) %>%
  filter(white_hisp_diff != "NA") %>%
  ggplot(aes(fct_reorder(county, white_hisp_diff), white_hisp_diff)) +
  geom_point(col = "darkred") +
  coord_flip() + 
  gghighlight::gghighlight(white_hisp_diff < -.25 |
                             white_hisp_diff > .25,
                           unhighlighted_colour = alpha("steelblue", 0.4)) + 
  geom_point(col = "darkred", size = 2.5) + 
  labs(title = "Differences Self-Regulation Scores by Race in Oregon",
       subtitle = "Disparities in scores between White and Hispanic/Latino students",
       x = NULL,
       y = "Difference in scores")

```

---
# Visualization 3: self-regulation scores over time (2014-2018)
```{r echo=FALSE, include=FALSE}

dfirstwave <- import(here("data", "ka_2013-14_lookback_report.xlsx")) %>%
  clean_names()

dfirstwave1 <- dfirstwave %>%
  rename(county = oregon_kindergarten_assessment,
         district_id = x2,
         district = x3,
         inst_id = x4,
         inst_name = x5,
         inst_type = x6,
         stu_grp_type = x7,
         stu_grp = x8,
         avg_self_reg = x9) %>%
  select(c(1:9))

dfirstwave1 <- dfirstwave1 %>%
  filter(avg_self_reg != "*")

dfirstwave1 = dfirstwave1[-1,]

dfirstwave1$avg_self_reg = as.numeric(dfirstwave1$avg_self_reg)

dfirstwave2 <- dfirstwave1 %>%
  group_by(county) %>%
  summarise(avg_self_reg_county_firstwave = mean(avg_self_reg))

dfirstwave2 <- dfirstwave2 %>%
  mutate(county = str_to_lower(county))

d_tidy <- d_tidy %>%
  mutate(county = str_to_lower(county))

join2 <- left_join(d_tidy, dfirstwave2)

join2 <- join2 %>%
  rename(`2014` = avg_self_reg_county_firstwave,
         `2018` = avg_self_reg)

join2$`2014` = as.numeric(join2$`2014`)
join2$`2018` = as.numeric(join2$`2018`)

join2 <- join2 %>%
  select(1, 9, 10)

join2 <- join2 %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year", 
    values_to = "Score",
  )

```
**OKA aims to:** 
1. identify systemic opportunity gaps
2. determine resource allocation to best support students in need 
3. measure improvement over time


**Purpose of Visualization 3:**

To address all of the OKA aims by demonstrating change in self-regulation scores over time by county. 

**Audience:** 

Oregon Department of Education, policymakers

---
### Viz 3: initial try  
```{r echo=FALSE, warning=FALSE}

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  ggplot(aes(county, avg_score)) +
  geom_line(aes(group = county), color = "gray40") +
  geom_point(aes(color = Year)) +
  coord_flip() +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       x = NULL,
       y = "Score (on 1-5 scale)")

```

---
### Viz 3: order according to average score, remove legend

```{r echo=FALSE, warning=FALSE}

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  filter(county != "all counties") %>%
  ungroup() %>%
  ggplot(aes(fct_reorder(county, avg_score), avg_score)) +
  geom_line(aes(group = county), color = "gray40") +
  geom_point(aes(color = Year), size = 1.5) +
  coord_flip() +
  theme(
    plot.title.position = "plot",
    plot.subtitle = element_markdown()) +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       subtitle = "<span style = 'color:#FF6C67;'>2014</span> compared to <span style = 'color:#00C2C6;'>2018 </span>change.",
       x = NULL,
       y = "Score (on 1-5 scale)") + 
  guides(color = "none")

```

---
### Viz 3: order according to *change* in score rather than *average* score
```{r echo=FALSE, warning=FALSE}

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "avg_score"
  ) %>%
  mutate(change_14_to_18 = c(`2018` - `2014`)) %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year",
    values_to = "Score"
    ) %>%
  ggplot(aes(fct_reorder(county, change_14_to_18), Score)) +
  geom_point(aes(color = Year), size = 1.5) +
  geom_line(aes(group = county), color = "gray40") +
  coord_flip() +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       x = NULL,
       y = "Score (on 1-5 scale)")

```


---
### Viz 3: further refining (color, geom sizes, theme, subtitle)

```{r echo=FALSE, warning=FALSE}

join2 <- join2 %>%
  mutate(county = str_to_title(county))

join2 %>%
  group_by(county, Year) %>%
  summarise(avg_score = mean(Score)) %>%
  pivot_wider(
    names_from = "Year",
    values_from = "avg_score"
  ) %>%
  mutate(change_14_to_18 = c(`2018` - `2014`)) %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year",
    values_to = "Score"
    ) %>%
  filter(county != "all counties") %>%
  ggplot(aes(fct_reorder(county, -change_14_to_18), Score)) +
  geom_line(aes(group = county), color = "gray40", size = 1) +
  geom_point(aes(color = Year), size = 2.2) +
  coord_flip() +
  theme_minimal() +
  labs(title = "Self Regulation Score Changes from 2014 to 2018",
       subtitle = "Ordered from greatest increase to greatest decrease",
       x = NULL,
       y = "Score (on 1-5 scale)") + 
  scale_color_brewer(palette = "Set1")

```

---
# Challenges along the way
### Mapping data onto aesthetics: thinking in layers

```{r eval=FALSE}

ggplot(aes(x,y)) %>%
  geom(aes(???))

```
### Adjusting proportions 
### Problem-solving via googling:
- Last term it never worked. 
- Now it works sometimes! Which is so exciting!! 
- But it's still hard to know how long to struggle with something...
---
# Other challenges: **very** ugly maps! 
```{r echo=FALSE, warning=FALSE}

join1 %>%
ggplot() + 
    geom_polygon(aes(long, lat, group = group, fill = avg_self_reg_county)) + 
    coord_fixed(1.3) +
    scale_fill_viridis_c(name = "Score (1-5 scale)", 
                         option = "inferno") +
    labs(title = "Average Self Regulation Scores by County in Oregon",
         x = NULL,
         y = NULL) +
     theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank()) +
  geom_text(data = or_county, aes(long, lat, label = county), size = 2)

```

---
# Successes along the way: 
### Pivoting and regular expression
```{r eval=FALSE}

join2 <- join2 %>%
  pivot_longer(
    cols = 2:3,
    names_to = "Year", 
    values_to = "Score",
*    names_pattern = ".+_.+_.+_(.+)"
  )
```
### Also, tidying -> joining datasets
---

class: center, middle

# Thank you!
